---
title: "Lab 5: Rejection Sampling"
author: "Rebecca C. Steorts"
date: "August 24, 2020"
output: 
     pdf_document:
      includes: 
          in_header: custom2.tex
---

# Agenda

We can often end up with posterior distributions that we only know up to a normalizing constant. For example, in practice, we may derive $$p(\theta \mid x) \propto p(x\mid \theta) p(\theta)$$ and find that the normalizing constant $p(x)$ is very difficult to evaluate. Such examples occur when we start building non-conjugate models in Bayesian statistics. 

Given such a posterior, how can we appropriate it's density? One way is using rejection sampling. As an example, let's suppose our resulting posterior distribution is  $$f(x) \propto sin^2(\pi x), x \in [0,1].$$

In order to understand how to approximate the density (normalized) of $f$, we will investigate the following tasks:


1. Plot the densities of $f(x)$ and the Unif(0,1) on the same plot. 
 According to the rejection sampling approach sample from $f(x)$ using the Unif(0,1) pdf as an enveloping function.
2.  Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.
3.  Repeat Tasks 1 - 3 for  Beta(2,2) as an enveloping function. 
4.  Provide the four histograms from Tasks 2 and 3 using the Uniform(0,1) and the Beta(2,2) enveloping proposals. Provide the acceptance ratios. Provide commentary. 
5.  Do you recommend the Uniform or the Beta(2,2) as a better enveloping function (or are they about the same)? If you were to try and find an enveloping function that had a high acceptance ratio, which one would you try and why? 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Task 1

Plot the densities of $f(x)$ and the Unif(0,1) on the same plot. 

Let's first create a sequence of points from 0 to 1, so that we can have a grid of points for plotting both of the proposed functions. 

```{r}
set.seed(1)
# grid of points
x <- seq(0, 1, 10^-2)
```

```{r}
fx <- function(x) sin(pi * x)^2

# plot f(x)
plot(fx, xlim = c(0,1), ylim = c(0,1.5), ylab = "f(x)", lwd = 2)

# add uniform distribution
curve(dunif, add = TRUE, col = "blue", lwd = 2) 

# add beta distribution
curve(dbeta(x,2,2), add = TRUE, col = "red", lwd = 2)

legend("bottom", legend = c(expression(paste("sin(",pi,"x)"^"2")),"Unif(0,1)",
"Beta(2,2)"), 
col = c("black", "blue", "red"), 
lty = c(1,1,1), 
bty = "n", 
cex = 1.1, 
lwd = 2)
```

Figure 1: Comparision of the target function and the Unif(0,1) and the Beta(2,2) densities on the same plot.

# Tasks 2 -- 4

According to the rejection sampling approach sample from f(x) using the Unif(0,1) pdf as an enveloping function. In order to do this, we write a general rejection sampling function that also allows us to plot the historams for any simulation size. Finally, our function also allows us to look at task 4 quite easily. 

```{r}
# f = sin
sim_fun <- function(f, # function of interest fx
                    envelope = "unif", # default = uniform
                    par1 = 0, par2 = 1, # parameters
                    n = 10^2, # simulation size
                    plot = TRUE){ # create plot
  
  # r: n samples - x1, x2, ..., xn
  # d: n samples - f(x1), f(x2), ..., f(xn)
  
  # paste0: create "runif" or "dbeta"
  # match.fun find the function named "runif" or "dbeta"
  r_envelope <- match.fun(paste0("r", envelope)) 
  d_envelope <- match.fun(paste0("d", envelope)) 
  
  # runif(n, par1, par2) 
  # dbeta(n, shape1, shape2)
  proposal <- r_envelope(n, par1, par2) 
  
  # the bigger the density ratio, the more likely to accept the sample
  # the better the enveloping function is
  # density_ratio < 1
  density_ratio <- f(proposal) / d_envelope(proposal, par1, par2) # f(x)/g(x)
  
  # if the density ratio is greater than a random draw from runif, accept the sample
  # compare density with uniform distribution
  # return Boolean: if True, keep in samples
  samples <- proposal[runif(n) < density_ratio] 
  
  acceptance_ratio <- length(samples) / n
  
  if (plot) {
    hist(samples, probability = TRUE, 
         main = paste0("Histogram of ", 
                       n, " samples from ", 
                       envelope, "(", par1, ",", par2,
                       ").\n Acceptance ratio: ",
                       round(acceptance_ratio, 2)), 
                       cex.main = 0.75)
  }
  
  list(x = samples, acceptance_ratio = acceptance_ratio)
}
```

```{r}
par(mfrow = c(2,2), mar = rep(4, 4))
unif_1 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^2) 
unif_2 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^5)
beta_1 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^2) 
beta_2 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^5)
```

Figure 2: Comparision of the output of the rejection sampling for 100 versus 100,000 simulations with uniform and beta distributions as envelope functions.

## Task 4
Provide the four histograms from Tasks 2 and 3 using the Uniform(0,1) and the
Beta(2,2) enveloping proposals. Provide the acceptance ratios. Provide commentary.

When n = 100, the acceptance ratio from unif(0, 0) is 0.48, and the acceptance 
ratio from beta(2, 2) is 0.41.
When n = 1e5, the acceptance ratio from unif(0, 0) is 0.5, and the acceptance 
ratio from beta(2, 2) is 0.5.

When the sample size is big enough, the acceptance ratio from both the uniform 
and beta are the same. However, when the sample size is small, the acceptance
ratio from beta is lower than the uniform. 


## Task 5
Do you recommend the Uniform or the Beta(2,2) as a better enveloping function (or
are they about the same)? If you were to try and find an enveloping function that
had a high acceptance ratio, which one would you try and why? 

It's hard to decide. First, when the sample size grows, both the uniform and Beta(2,2)has acceptance ratio of 0.5. Also, towards the center, the shapes of f(x) and beta(2,2) don't align well, so the uniform tends to have a higher density
ratio than beta(2,2) in that region. While on the 2 sides, the shapes of f(x) and uniform don't align well, so the beta(2,2) tends to have a higher density ratio 
than the uniform in that region.

I would try a distribution that has similar shape to f(x), like a normal distribution with mean at 0.5.